Detection and replication of epistasis influencing transcription in humans

### 2013-09-01 docs/manuscript/methods.tex
most recent version

\begin{equation}
y = \mu + \beta_{1}c + \beta_{2}p + \beta_{3}s + \beta_{4}a + g + e
\label{eq:lm}
\end{equation}
where $\mu$ is the population mean expression levels, $c$, $p$, $s$ and $a$ are vectors of chip, chip position, sex and generation respectively, fitted as fixed effects; and $g$ is a random additive polygenic effect with a variance covariance matrix 
\begin{equation}
G_{ijk} = \left \{ 
\begin{array}{ll}
\sigma _a ^2 + \sigma _e ^2&        j = k \\ 
2\phi _{ijk} \sigma _a ^2& 			j \neq k \\
\end{array} \right.
\end{equation}
The parameters $\sigma_a^2$ and $\sigma_e^2$ are variance components for additive background genetic and environmental effects respectively. Here, we are using family based pedigree information rather than SNP based IBD to account for relationships between individuals and so $\phi _{ijk}$ is the kinship coefficient between individuals $j$ and $k$. The residual, $e$, from equation \ref{eq:lm} is assumed to follow a multivariate normal distribution with a mean of zero. Residuals were normalised by rank transformation and used as the adjusted phenotype for the pairwise epistasis scan to remove any skewness and avoid results being driven by outliers. The GenABEL package for R was used to perform the normalisation \cite{Aulchenko2007}.

@@

@@

Following filtering on this threshold only SNP pairs with at least 5 data points in all 9 genotype classes were kept. We then calculated the LD between interacting SNPs (amongst unrelated individuals within the discovery sample and also from 1000 genomes data) and removed any pairs with $r^2 > 0.1$ or $D'^{2} > 0.1$ to avoid the inclusion of haplotype effects and to increase the accuracy of genetic variance decomposition. If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinel SNP pair was retained, \emph{i.e.} if a probe had multiple SNP pairs that were on chromosomes one and two then only the SNP pair with the most significant $p$-value was retained. At this stage 6404 filtered SNP pairs remained. We then performed a second filtering screen that was identical to the first but an additional step was included where any SNPs that had previously been shown to have a significant additive or dominant effect ($p < 1.29 \times 10^{-11}$) was removed, creating a second set of 4751 filtered SNP pairs.

@@

@@

>>>>

\begin{equation}
y = \mu + \beta_{1}c + \beta_{2}p + \beta_{3}s + \beta_{4}a + g + e
\label{eq:lm}
\end{equation}
where $\mu$ is the population mean expression levels, $c$, $p$, $s$ and $a$ are vectors of chip, chip position, sex and generation respectively, fitted as fixed effects; and $g$ is a random additive polygenic effect with a variance covariance matrix 
\begin{equation}
G_{jk} = \left \{ 
\begin{array}{ll}
\sigma _a ^2             &          j = k \\ 
2\phi _{ijk} \sigma _a ^2& 			j \neq k \\
\end{array} \right.
\end{equation}
The parameter $\sigma_a^2$ is the variance component for additive background genetic. Here, we are using family based pedigree information rather than SNP based IBD to account for relationships between individuals and so $\phi _{ijk}$ is the kinship coefficient between individuals $j$ and $k$. The residual, $e$, from equation \ref{eq:lm} is assumed to follow a multivariate normal distribution with a mean of zero. Residuals were normalised by rank transformation and used as the adjusted phenotype for the pairwise epistasis scan to remove any skewness and avoid results being driven by outliers. The GenABEL package for R was used to perform the normalisation \cite{Aulchenko2007}.

@@

\subsection{Two stage search}

@@

Following filtering on this threshold only SNP pairs with at least 5 data points in all 9 genotype classes were kept. We then calculated the LD between interacting SNPs (amongst unrelated individuals within the discovery sample and also from 1000 genomes data) and removed any pairs with $r^2 > 0.1$ or $D'^{2} > 0.1$ to avoid the inclusion of haplotype effects and to increase the accuracy of genetic variance decomposition. If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinel SNP pair was retained, \emph{i.e.} if a probe had multiple SNP pairs that were on chromosomes one and two then only the SNP pair with the most significant $p$-value was retained. At this stage 6404 filtered SNP pairs remained (filter 1). We then performed a second filtering screen that was identical to the first but an additional step was included where any SNPs that had previously been shown to have a significant additive or dominant effect ($p < 1.29 \times 10^{-11}$) was removed, creating a second set of 4751 filtered SNP pairs (filter 2).

@@

\subsection{Type 1 error rate}

Using a Bonferroni correction of $0.05$ in the second stage of the two stage discovery scan implies a type 1 error rate of $\alpha = 0.05$. However, this could be underestimated because the number tests performed in the second stage depends on the number of tests in the first stage, and this depends on statistical power and model choice. We performed simulations to estimate the type 1 error rate of this study design.

We assumed a null model where there was one true additive effect and 7 other terms with no effect. To simulate a test statistic we simulated 8 z-scores, $z_1 \sim N(\sqrt{NCP}, 1)$ and $z_{2..8} \sim N(0,1)$. Thus $z_{full} = \sum^8_{i=1}{z_i} \sim \chi^{2}_{8}$ (representing the 8 d.f. test) and $z_{int} = \sum^8_{i=5}{z_i} \sim \chi^{2}_{4}$ (representing the 4 d.f. test where the null hypothesis of no epistasis is true). For a particular value of $NCP$ we simulated 100,000 $z$ values, and calculated the $p_{full}$-value for the $z_{full}$ test statistic. The $n_{int}$ test statistics with $p_{full} < 2.31 \times 10^{-16}$ were kept for the second stage, where the type 1 error rate of stage 2 was calculated as the proportion of $p_{int} < 0.05 / n_{int}$. The power at stage 1 was calculated as $n_{int} / 100,000$. This procedure was performed for a range of $NCP$ parameters that represented power ranging from $\sim 0$ to $\sim 1$.

@@

\subsection{Concordance of direction of effects}

We used four methods to calculate the concordance of the direction of effects between the discovery and replication datasets.

\begin{enumerate}

\item [Test 1] Is the most significant epistatic effect in the discovery set in the same direction as the same epistatic effect in the replication sets? We decomposed the genetic variance into 8 orthogonal effects, four of which are epistatic ($A \times A$, $A \times D$, $D \times A$, $D \times D$). The sign of the epistatic effect that had the largest variance in the discovery was recorded, and then was compared to the same epistatic effect in the two replication datasets (regardless of whether or not the same epistatic effect was the largest in the replication datasets). The probability of the sign being the same in one dataset is $1/2$. The probability of the sign being the same in two is $1/4$.

\item [Test 2] Is the most significant epistatic effect in the discovery the same as the largest epistatic effect in the replication set with the sign being concordant. As in Test 1, but this time we required that the largest effect was the same in the discovery and the replication, and that they had the same sign (\emph{e.g.} if the largest effect in the discovery is $A \times A$, with a positive effect, then concordance is achieved if the same is true in the replication). The probability of one replication dataset being concordant by chance is $1/8$, and concordance in both is $1/64$.

\item [Test 3] Are all 4 epistatic effects in the same direction in the discovery and the replication datasets? This is a very conservative test as it is not realistic to assume that all four epistatic effects are significant in every interaction. The probability of concordance in one replication dataset is $1/16$, and the probability of concordance in both replication datasets is $1/256$.

\item [Test 4] Do the epistatic effects that are significant at nominal $p < 0.05$ in the discovery have the same direction of effect as in the replication? Here we count all the epistatic variance components in the discovery that have $p < 0.05$ (1133 amongst the 434 discovery SNP pairs, $i.e.$ each SNP pair has at least 1 and at most 4 significant epistatic variance components). Then we compare the direction of the effect in the replication dataset. The probability of the sign being the same in one dataset for any one significant effect is $1/2$. The probability of the sign being the same in two is $1/4$.

\end{enumerate}

The probability of observing the number of concordant signs is calculated using a binomial test.

### 2013-09-01 docs/manuscript/nature_letter.tex
most recent version

\author[1,2,*]{Gibran Hemani}
\author[1,2]{Konstantin Shakhbazov}
\author[3]{Harm-Jan Westra}
\author[4,5,6]{Tonu Esko}
\author[7]{Anjali K Henders}
\author[1,2]{Allan F. McRae}
\author[2]{Jian Yang}
\author[8]{Greg Gibson}
\author[7]{Nick G Martin}
\author[4]{Andres Metspalu}
\author[3]{Lude Franke}
\author[7,+]{Grant W Montgomery}
\author[1,2,+]{Peter M Visscher}
\author[1,2,+]{Joseph E Powell}

@@

Epistasis is the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome. The extent to which epistasis influences complex traits \cite{Carlborg2004} and contributes to their variation \cite{Hill2008a, Crow2010} is a fundamental question in evolution and human genetics. Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and some examples have been reported in other species \cite{Carlborg2006}, few convincing examples exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}. Its absence from empirical findings may simply be due to its low incidence in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but an alternative view is that it has previously been too technically challenging to detect due to statistical power and computational issues \cite{Cordell2009}. Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design, many instances of epistasis are found between common single nucleotide polymorphisms (SNPs). In a cohort of 846 individuals with data on 7339 gene expression levels in peripheral blood, we found 501 significant pairwise epistatic interactions between common SNPs acting on the expression levels of 238 genes ($p < 2.91 \times 10^{-16}$). We tested the discovery interactions for replication in two independent data sets \cite{Metspalu2004, Fehrmann2011}. Three hundred and forty-five interactions had replication interaction $p$-values that were more extreme than the 2.5\% confidence interval of the distribution under the null hypothesis of no epistasis, with 30 significant at a conservative $p < 0.05$ Bonferroni level. There was evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 2Mb of regions of known physical chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$). Epistatic networks of three SNPs or more influence the expression levels of 129 genes, whereby one \emph{cis}-acting SNP is modulated by several \emph{trans}-acting SNPs. For example MBNL1 is influenced by an additive effect at rs13069559 which itself is masked by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction. This study presents the first evidence for multiple instances of epistatic genetic effects emerging from natural genetic variation in humans.

@@

% \subsection{Introduction}
In the genetic analysis of complex traits it is usual for SNP effects to be estimated using an additive model where they are assumed to contribute independently and cumulatively to the mean of a trait. This framework has been successful in identifying thousands of associations \cite{Visscher2012}. But to date, though its contribution to phenotypic variance is frequently the subject of debate \cite{Carlborg2004, Hill2008a, Crow2010}, there is little empirical exploration of the role that epistasis plays in the architecture of complex traits in humans \cite{Strange2010, Evans2011}. Outside the prism of human association studies there is evidence for epistasis, not only at the molecular scale from artificially induced mutations \cite{Costanzo2010} but also at the evolutionary scale in fitness adaptation \cite{Weinreich2006} and speciation \cite{Breen2012}.

Methods are now available to overcome the computational problems involved in searching for epistasis, but its detection still remains problematic due to reduced statistical power. For example increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs \cite{Weir2008, Hemani2013}, increased model complexity in fitting interaction terms \cite{Marchini2005}, and more extreme significance thresholds to account for increased multiple testing \cite{Cordell2009} all make it more difficult to detect epistasis in comparison to additive effects. Thus, when combined with small genetic effect sizes, as is expected in most complex traits of interest \cite{Visscher2012}, the power to detect epistasis diminishes rapidly. There are two simple ways to overcome this problem. One is by using extremely large sample sizes \cite{LangoAllen2010}; another is by analysing traits that are likely to have large effect sizes among common variants. Because our focus was to ascertain the extent to which instances of epistasis occur amongst natural genetic variation we designed a study around the latter approach and searched for epistatic genetic effects that influence gene expression levels. Transcription levels can be measured for thousands of genes. These traits are largely heritable but on average less polygenic than high level phenotypes \cite{Powell2013}, thus many genetic effects are relatively large, maximising the chance at detecting epistasis, should it exist.

% \subsection{Initial search in discovery set and replication}
In our discovery dataset (Brisbane Systems Genetics Study, BSGS \cite{Powell2012}) of 846 individuals genotyped at 528,509 SNPs, we exhaustively tested every pair of SNPs for genetic interactions against each of 7339 expression traits in peripheral blood. After stringent filtering and multiple testing correction (5\% significance threshold $ p < 2.91 \times 10^{-16}$, Methods) we identified 501 putative genetic interactions influencing the expression levels of 238 genes (Supplementary Table \ref{tab:results}). Of the 501 discovery interactions, 434 had available data and passed filtering (Methods) in two independent replication datasets, Fehrmann \cite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \cite{Metspalu2004}, in which we saw convincing evidence for replication. We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 30 were significant after applying a Bonferroni correction for multiple testing (5\% significance threshold $p < 0.0001$, Table \ref{tab:bonferroni}). These significant interactions exhibited remarkable similarity in GP maps between all three datasets (Figure \ref{fig:gpmaps}).

In addition, using the meta analysis from the replication samples only, we observed that 316 of the remaining 404 discovery SNPs had replication interaction $p$-values more extreme than the 2.5\% confidence interval of the distribution under the null distribution of no epistatic effects ($p << 1.0 \times 10^{-16}$, Figure \ref{fig:qqMeta} and Supplementary Figure \ref{fig:qqMetaNonsig}). The congruence of the epistatic networks in discovery and replication datasets is shown in Figure \ref{fig:fireworks}, demonstrating that these complex genetic patterns are common even across independent datasets. A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 20 of the SNP pairs passed filtering because the sample size was small ($n=139$), and likely due to insufficient power we found no evidence for replication (Supplementary Figure \ref{fig:qqgg}). It should be noted that although it is a necessary step to establish the veracity of the signals from the discovery set, replication of epistasis is difficult in practice because the dependence on LD between observed SNPs and causal variants is up to three orders of magnitude higher than it is for independent additive effects \cite{Weir2008, Hemani2013}. Therefore these results are encouraging with regards to the detection and replication of epistasis.

% \subsection{Patterns of epistasis}
Though seldom the focus of association studies, SNPs with known main effects are often tested for additive $\times$ additive genetic interactions \cite{Cordell2009}, but our analysis shows that this is unlikely to be the most effective strategy for its detection. The majority of our discovery interactions comprised of one SNP that was significantly associated with the gene expression level in the discovery dataset, and one SNP that had no previous association \cite{Powell2013} (439 out of 501, Methods). Only nine interactions were between SNPs that both had known main effects while 64 were between SNPs that had no known main effects. Additionally, we observed that the largest epistatic variance component for the 501 interactions was equally divided amongst additive $\times$ additive, additive $\times$ dominance, dominance $\times$ additive and dominance $\times$ dominance at the discovery stage ($p = 0.22$ for departure from expectation). This is not surprising because the patterns of epistasis used for statistical decomposition are not designed to resemble biological function \cite{Cockerham1954}.

@@

We also demonstrate physical organisation of interacting loci within the cell, suggesting a mechanism by which biological function can lead to epistatic genetic variance. It has been shown that different chromosomal regions spatially colocalise in the cell through chromatin interactions \cite{Lieberman-Aiden2009}. We cross-referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96,139$) in K562 blood cell lines \cite{Lan2012} (Methods) and found that 44 epistatic interactions mapped to within 2Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}). Interaction of distant loci may occur through physical proximity in transcriptional factories that organise across different chromosome regions and can regulate transcription of related genes \cite{Osborne2004, Rieder2012}.

% \subsection{Contribution of epistatic variance}
Though we present many instances of epistasis, quantifying its relative importance to complex traits in humans remains an open question. In this study we are able to identify 238 gene expression traits with at least one significant interaction given our experiment-wide threshold. How does this compare to the number of traits influenced by additive effects? The BSGS dataset has been previously analysed for additive effects at all expression traits \cite{Powell2012}, and if we take all the additive eQTLs that were significant at the epistatic threshold of $p < 2.91 \times 10^{-16}$ we find that 453 gene expression levels out of the 7339 analysed had at least one significant expression quantitative trait locus (eQTL). Therefore it can be argued that the number of instances of detectable epistasis are substantial.

@@

@@

\begin{ThreePartTable}
\label{tab:results}
\begin{longtable}{|llr|lrrl|lrrl|rrrr|r|}

\caption{\textbf{Details on 501 interactions discovered in BSGS dataset}}\\

@@

>>>>

\author[1,2,*]{Gibran Hemani}
\author[1,2]{Konstantin Shakhbazov}
\author[3]{Harm-Jan Westra}
\author[4,5,6]{Tonu Esko}
\author[7]{Anjali K Henders}
\author[1,2]{Allan F. McRae}
\author[2]{Jian Yang}
\author[8]{Greg Gibson}
\author[7]{Nicholas G Martin}
\author[4]{Andres Metspalu}
\author[3]{Lude Franke}
\author[7,+]{Grant W Montgomery}
\author[1,2,+]{Peter M Visscher}
\author[1,2,+]{Joseph E Powell}

@@

Epistasis is the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome. The extent to which epistasis influences complex traits \cite{Carlborg2004} and contributes to their variation \cite{Hill2008a, Crow2010} is a fundamental question in evolution and human genetics. Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and examples have been reported in other species \cite{Carlborg2006}, few convincing examples with independent replication exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}. Its absence from empirical findings may simply be due to its low incidence in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but an alternative view is that it has previously been too technically challenging to detect due to statistical power and computational issues \cite{Cordell2009}. Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design, many instances of epistasis are found between common single nucleotide polymorphisms (SNPs). In a cohort of 846 individuals with data on 7339 gene expression levels in peripheral blood, we found 501 significant pairwise epistatic interactions between common SNPs acting on the expression levels of 238 genes ($p < 2.91 \times 10^{-16}$). Replication of these signals in two independent data sets \cite{Metspalu2004, Fehrmann2011} showed both concordance of direction of epistatic effects ($p = 5.56 \times 10^{-31})$ and enrichment of interaction $p$-values, with 30 being significant at a conservative threshold of $p < 0.05 / 434$. There was evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 5Mb of regions of known physical chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$). Epistatic networks of three SNPs or more influence the expression levels of 129 genes, whereby one \emph{cis}-acting SNP is modulated by several \emph{trans}-acting SNPs. For example MBNL1 is influenced by an additive effect at rs13069559 which itself is masked by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction. This study presents the first evidence for multiple instances of natural genetic polymorphisms interacting to influence human traits.

@@

% \subsection{Introduction}
In the genetic analysis of complex traits it is usual for SNP effects to be estimated using an additive model where they are assumed to contribute independently and cumulatively to the mean of a trait. This framework has been successful in identifying thousands of associations \cite{Visscher2012}. But to date, though its contribution to phenotypic variance is frequently the subject of debate \cite{Carlborg2004, Hill2008a, Crow2010}, there is little empirical exploration of the role that epistasis plays in the architecture of complex traits in humans \cite{Strange2010, Evans2011}. Beyond the prism of human association studies there is evidence for epistasis, not only at the molecular scale from artificially induced mutations \cite{Costanzo2010} but also at the evolutionary scale in fitness adaptation \cite{Weinreich2006} and speciation \cite{Breen2012}.

Methods are now available to overcome the computational problems involved in searching for epistasis, but its detection still remains problematic due to reduced statistical power. For example increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs \cite{Weir2008, Hemani2013}, increased model complexity in fitting interaction terms \cite{Marchini2005}, and more extreme significance thresholds to account for increased multiple testing \cite{Cordell2009} all make it more difficult to detect epistasis in comparison to additive effects. Thus, when combined with small genetic effect sizes, as is expected in most complex traits of interest \cite{Visscher2012}, the power to detect epistasis diminishes rapidly. There are two simple ways to overcome this problem. One is by using extremely large sample sizes \cite{LangoAllen2010}; another is by analysing traits that are likely to have large effect sizes among common variants. Because our focus was to ascertain the extent to which instances of epistasis occur amongst natural genetic variation we designed a study around the latter approach and searched for epistatic genetic effects that influence gene expression levels. Transcription levels can be measured for thousands of genes. These traits are largely heritable but on average less polygenic than high level phenotypes, thus many genetic effects are relatively large \cite{Powell2013}, maximising the chance at detecting epistasis, should it exist.

% \subsection{Initial search in discovery set and replication}
In our discovery dataset (Brisbane Systems Genetics Study, BSGS \cite{Powell2012}) of 846 individuals genotyped at 528,509 SNPs, we used a two stage approach to identify genetic interactions. First, we exhaustively test every pair of SNPs for pairwise effects against each of 7339 expression traits in peripheral blood (5\% significance threshold $ p < 2.91 \times 10^{-16}$, Methods). Second, we filtered the SNP pairs from stage 1 on LD and genotype class counts, and tested the remaining pairwise effects for significant interaction terms and used a Bonferroni correction for multiple testing (estimated type 1 error rate $\alpha \approx 0.07$, Methods, Supplementary Figure \ref{fig:type1}). Using this design we identified 501 putative genetic interactions influencing the expression levels of 238 genes (Supplementary Table \ref{tab:results}). Of the 501 discovery interactions, 434 had available data and passed filtering (Methods) in two independent replication datasets, Fehrmann \cite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \cite{Metspalu2004}, in which we saw convincing evidence for replication. We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 30 were significant after applying a Bonferroni correction for multiple testing (5\% significance threshold $p < 0.05/434$, Table \ref{tab:bonferroni}). These significant interactions exhibited remarkable similarity in GP maps between all three datasets (Figure \ref{fig:gpmaps}). We decomposed the genetic effects of each of the SNP pairs into orthogonal additive, dominance and epistatic effects ($A1$, $A2$, $D1$, $D2$, $A \times A$, $A \times D$, $D \times A$, $D \times D$) and tested for concordance of the sign of the most signicant effect in the independent datasets (Methods). The most significant epistatic effect in the discovery sample was in the same direction in both replication datasets in 22 out of the 30 significantly replicated interactions ($p = 3.76 \times 10^{-8}$, Supplementary Table \ref{tab:sign}).

In addition, using the meta analysis from the replication samples only, we observed that 316 of the remaining 404 discovery SNP pairs had replication interaction $p$-values more extreme than the 2.5\% confidence interval of the quantile-quantile plot against the null hypothesis of no interactions ($p << 1.0 \times 10^{-16}$, Figure \ref{fig:qqMeta} and Supplementary Figure \ref{fig:qqMetaNonsig}). Concordance of the direction of the effect of the largest variance component was also highly significant ($p = 5.71 \times 10^{-31}$, Supplementary Table \ref{tab:sign}). The congruence of the epistatic networks in discovery and replication datasets is shown in Figure \ref{fig:fireworks}, demonstrating that these complex genetic patterns are common even across independent datasets. A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 20 of the SNP pairs passed filtering because the sample size was small ($n=139$), and likely due to insufficient power we found no evidence for replication (Supplementary Figure \ref{fig:qqgg}). It should be noted that although it is a necessary step to establish the veracity of the signals from the discovery set, replication of epistasis is difficult in practice because the dependence on LD between observed SNPs and causal variants is up to three orders of magnitude higher than it is for independent additive effects \cite{Weir2008, Hemani2013}. Therefore these results are encouraging with regards to the detection and replication of epistasis.

% \subsection{Patterns of epistasis}
Though seldom the focus of association studies, SNPs with known main effects are often tested for additive $\times$ additive genetic interactions \cite{Cordell2009}, but our analysis shows that this is unlikely to be the most effective strategy for its detection. The majority of our discovery interactions comprised of one SNP that was significantly associated with the gene expression level in the discovery dataset, and one SNP that had no previous association \cite{Powell2013} (439 out of 501, Methods). Only nine interactions were between SNPs that both had known main effects while 64 were between SNPs that had no known main effects. Additionally, we observed that the largest epistatic variance component for the 501 interactions was equally divided amongst additive $\times$ additive, additive $\times$ dominance, dominance $\times$ additive and dominance $\times$ dominance at the discovery stage ($p = 0.22$ for departure from expectation). This is not surprising because the patterns of epistasis used for statistical decomposition (\emph{i.e.} $A \times A$, $A \times D$, $D \times A$, $D \times D$) are simply convenient orthogonal parameterisations of a two locus model, and are not intended to model biological function \cite{Cockerham1954}.

@@

We also demonstrate physical organisation of interacting loci within the cell, suggesting a mechanism by which biological function can lead to epistatic genetic variance. It has been shown that different chromosomal regions spatially colocalise in the cell through chromatin interactions \cite{Lieberman-Aiden2009}. We cross-referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96,139$) in K562 blood cell lines \cite{Lan2012} (Methods) and found that 44 epistatic interactions mapped to within 5Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}). Interaction of distant loci may occur through physical proximity in transcriptional factories that organise across different chromosome regions and can regulate transcription of related genes \cite{Osborne2004, Rieder2012}.

% \subsection{Contribution of epistatic variance}
Though we present many instances of epistasis, quantifying its relative importance to complex traits in humans remains an open question. In this study we are able to identify 238 gene expression traits with at least one significant interaction given our experiment-wide threshold. How does this compare to the number of traits influenced by additive effects? The BSGS dataset has been previously analysed for additive effects at all expression traits \cite{Powell2012}, and if we take all the additive eQTLs that were significant at the epistatic threshold of $p < 2.91 \times 10^{-16}$ we find that 453 gene expression levels out of the 7339 analysed had at least one significant expression quantitative trait locus (eQTL). Therefore it can be argued that the number of instances of detectable epistasis is substantial.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=5in]{type1}
	\caption{\textbf{Type 1 error rate of two stage design} In stage 1 SNPs are tested for full genetic effects (8 d.f.) and those that surpass a threshold for multiple testing are then tested for significant interaction terms in stage 2. These interaction $p$-values are then adjusted (Bonferroni) for the total number of tests that passed stage 1. The type 1 error rate of this two stage design is dependent on the power, which is not known empirically.}
\label{fig:type1}
\end{figure}
\clearpage

@@

\begin{ThreePartTable}
\begin{longtable}{|llr|lrrl|lrrl|rrrr|r|}

\caption{\textbf{Details on 501 interactions discovered in BSGS dataset}}  \label{tab:results} \\

@@

\begin{table}[ht]
{\footnotesize
\begin{threeparttable}
\caption{\textbf{Details on linkage disequilibrium and relative positions of all discovery \emph{cis-cis} interactions}}
\label{tab:ld}
\begin{tabular}{llllrrrrr}

  \hline
Chr & Gene & SNP 1 & SNP 2 & Position 1 & Position 2 & Distance / Mb & $R^2$ & $D'$ \\
  \hline
19 & TMEM149 & rs807491 & rs7254601 & 36268923 & 36147315.0 & 0.122 & 0.000 & 0.001 \\
  17 & FN3KRP & rs898095 & rs9892064 & 80890638 & 80827903.0 & 0.063 & 0.063 & 0.088 \\
  21 & CSTB & rs9979356 & rs3761385 & 45230974 & 45198355.0 & 0.033 & 0.041 & 0.066 \\
  3 & MBNL1 & rs16864367 & rs13079208 & 152234166 & 152116652.0 & 0.118 & 0.041 & 0.117 \\
  10 & ADK & rs2395095 & rs10824092 & 76446305 & 75929517.0 & 0.517 & 0.013 & 0.020 \\
  11 & CTSC & rs7930237 & rs556895 & 88117962 & 88077479.0 & 0.040 & 0.012 & 0.045 \\
  17 & GAA & rs11150847 & rs12602462 & 78153130 & 78146016.0 & 0.007 & 0.000 & 0.001 \\
  8 & NAPRT1 & rs2123758 & rs3889129 & 144663661 & 144613680.0 & 0.050 & 0.053 & 0.060 \\
  1 & LAX1 & rs1891432 & rs10900520 & 203877662 & 203780591.0 & 0.097 & 0.065 & 0.106 \\
  18 & MBP & rs8092433 & rs4890876 & 74747424 & 74732087.0 & 0.015 & 0.035 & 0.053 \\
  11 & SNORD14A & rs2634462 & rs6486334 & 17339127 & 17015557.0 & 0.324 & 0.008 & 0.012 \\
  21 & C21ORF57 & rs9978658 & rs11701361 & 48027084 & 47764477.0 & 0.263 & 0.032 & 0.065 \\
  16 & RPL13 & rs352935 & rs2965817 & 89648580 & 89513234.0 & 0.135 & 0.054 & 0.060 \\
  19 & ATP13A1 & rs4284750 & rs873870 & 19810050 & 19738554.0 & 0.071 & 0.008 & 0.015 \\
  2 & NCL & rs7563453 & rs4973397 & 232301670 & 232291471.0 & 0.010 & 0.027 & 0.029 \\
  5 & HNRPH1 & rs6894268 & rs4700810 & 179032488 & 178991794.0 & 0.041 & 0.000 & 0.001 \\
  19 & VASP & rs1264226 & rs2276470 & 46063167 & 45974668.0 & 0.088 & 0.018 & 0.022 \\
  7 & TRA2A & rs7776572 & rs11770192 & 23528927 & 23498358.0 & 0.031 & 0.064 & 0.064 \\
  21 & PRMT2 & rs2839372 & rs11701058 & 48063862 & 47776382.0 & 0.287 & 0.100 & 0.122 \\
  12 & OAS1 & rs13311 & rs2072133 & 113448652 & 113409260.0 & 0.039 & 0.002 & 0.016 \\
  16 & N4BP1 & rs12444224 & rs11649236 & 87580855 & 48632478.0 & 38.948 & 0.007 & 0.021 \\
  5 & CAST & rs12719343 & rs7733671 & 125369113 & 96000269.0 & 29.369 & 0.001 & 0.001 \\
  7 & DNAJB6 & rs2286842 & rs3779589 & 157216093 & 157163614.0 & 0.052 & 0.005 & 0.006 \\
  1 & OVGP1 & rs10802822 & rs1264898 & 240132968 & 111992823.0 & 128.140 & 0.008 & 0.030 \\
  20 & CD93 & rs2868504 & rs1884655 & 37771578 & 23074375.0 & 14.697 & 0.000 & 0.002 \\
  11 & PHCA & rs493642 & rs10736812 & 123097386 & 76708086.0 & 46.389 & 0.002 & 0.008 \\
  21 & MX1 & rs459498 & rs8130120 & 42795027 & 29363604.0 & 13.431 & 0.000 & 0.000 \\
  16 & AKTIP & rs2896940 & rs13332406 & 57721127 & 53489705.0 & 4.231 & 0.000 & 0.001 \\
  17 & CDK5R1 & rs9905940 & rs11655031 & 46614102 & 30833162.0 & 15.781 & 0.000 & 0.000 \\
  2 & CYBRD1 & rs888427 & rs7591849 & 172368120 & 160112881.0 & 12.255 & 0.000 & 0.000 \\
  8 & HMBOX1 & rs587639 & rs7837237 & 132725731 & 28876221.0 & 103.850 & 0.001 & 0.001 \\
  11 & TRAPPC4 & rs1793823 & rs3916581 & 131018917 & 118887887.0 & 12.131 & 0.001 & 0.002 \\
  12 & PEX5 & rs10444467 & rs4329748 & 128052636 & 7364442.0 & 120.688 & 0.000 & 0.000 \\
  12 & FLJ20489 & rs17615703 & rs3782908 & 117036766 & 48169526.0 & 68.867 & 0.001 & 0.002 \\
  16 & PRKCB1 & rs2188355 & rs10492793 & 23867776 & 12639800.0 & 11.228 & 0.000 & 0.000 \\
  14 & MRPL52 & rs1950857 & rs3811188 & 26710271 & 23299135.0 & 3.411 & 0.002 & 0.004 \\
  17 & C17ORF60 & rs9907897 & rs7405659 & 63502633 & 59874129.0 & 3.629 & 0.004 & 0.011 \\
  6 & FLJ43093 & rs6906101 & rs13214069 & 36667610 & 32705248.0 & 3.962 & 0.000 & 0.000 \\
  19 & TRAPPC5 & rs17159840 & rs17763599 & 7758194 & 2369415.0 & 5.389 & 0.000 & 0.000 \\
  22 & PISD & rs715572 & rs6518754 & 33234931 & 32097775.0 & 1.137 & 0.001 & 0.003 \\
  12 & DIP2B & rs871257 & rs12427378 & 117994348 & 51074199.0 & 66.920 & 0.001 & 0.001 \\
  12 & GPR162 & rs2272500 & rs2707210 & 79685913 & 6902002.0 & 72.784 & 0.003 & 0.005 \\
  17 & USP36 & rs2279308 & rs7225546 & 76794981 & 75151717.0 & 1.643 & 0.000 & 0.000 \\
   \hline
\end{tabular}
\end{threeparttable}
}
\end{table}
